---
title: "Smart Urban Parking Intelligence — R Analysis Report"
author: "Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 4.5)

# NOTE: knit this file with the PROJECT ROOT as the working directory, e.g.
#   rmarkdown::render("r-analytics/report/bda_analysis_report.Rmd", knit_root_dir = "../..")
# or, in RStudio, open the project at the repo root and knit from there.

suppressMessages({
  library(dplyr)
  library(ggplot2)
  library(lubridate)
  library(randomForest)
})
```

## 1. Introduction

Urban parking systems often lack real-time visibility into occupancy and near-term demand, which sends drivers circling for spaces and leaves operators unable to balance load across zones. This report works through the historical dataset produced by the project's data-generation pipeline (`scripts/generate_dataset.py`) using R: exploratory analysis, peak-hour and zone-utilization statistics, and a Random Forest model that predicts a zone's occupancy 30 minutes ahead.

The raw event stream, Kafka ingestion, and PySpark Structured Streaming job are implemented in Python (see the main `README.md`); this report and the accompanying `r-analytics/` scripts cover the statistical analysis, modeling, and dashboard side of the project in R.

## 2. Data Overview

```{r load-data}
raw <- read.csv("data/raw/parking_events.csv", stringsAsFactors = FALSE)
raw$timestamp <- as.POSIXct(raw$timestamp, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")

glimpse(raw)
```

```{r data-summary}
cat(sprintf("Total events : %d\n", nrow(raw)))
cat(sprintf("Zones        : %d\n", n_distinct(raw$zone_id)))
cat(sprintf("Vehicles     : %d\n", n_distinct(raw$vehicle_id)))
cat(sprintf("Date range   : %s to %s\n", min(raw$timestamp), max(raw$timestamp)))

table(raw$event_type)
```

```{r data-quality}
cat(sprintf("Rows with occupancy > capacity : %d\n", sum(raw$occupancy > raw$capacity)))
cat(sprintf("Rows with negative duration    : %d\n", sum(raw$parking_duration < 0)))
cat(sprintf("Duplicate event_id rows        : %d\n", sum(duplicated(raw$event_id))))
```

The dataset is clean by construction (the generator enforces `0 <= occupancy <= capacity` and non-negative durations), but a real deployment would need exactly these checks as a first-line data-quality gate — which is what `analytics/occupancy_analysis.py` and `r-analytics/occupancy_analysis.R` both do before any aggregation.

## 3. Occupancy & Utilization Analysis

```{r zone-utilization}
raw$utilization <- raw$occupancy / raw$capacity * 100

zone_util <- raw %>%
  group_by(zone_id) %>%
  summarise(avg_utilization = mean(utilization), max_utilization = max(utilization), .groups = "drop") %>%
  arrange(desc(avg_utilization))

zone_util
```

```{r zone-utilization-plot, fig.cap="Average utilization by zone"}
ggplot(zone_util, aes(x = reorder(zone_id, avg_utilization), y = avg_utilization)) +
  geom_col(fill = "#2c7fb8") +
  coord_flip() +
  labs(title = "Average Utilization by Zone", x = "Zone", y = "Average Utilization (%)") +
  theme_minimal()
```

## 4. Peak Hour Analysis

```{r peak-hours}
raw$hour <- hour(raw$timestamp)

hourly <- raw %>%
  group_by(hour) %>%
  summarise(avg_occupancy = mean(occupancy), .groups = "drop")

hourly %>% arrange(desc(avg_occupancy)) %>% head(5)
```

```{r peak-hours-plot, fig.cap="Average occupancy by hour of day"}
ggplot(hourly, aes(x = hour, y = avg_occupancy)) +
  geom_col(fill = "#f03b20") +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "Average Occupancy by Hour of Day", x = "Hour", y = "Average Occupancy") +
  theme_minimal()
```

Morning (8–10), lunch (12–13), and evening (17–19) rush periods stand out clearly, consistent with how the demand curve was calibrated in `scripts/generate_dataset.py`. On a real deployment this chart is exactly what you'd use to justify dynamic pricing windows or targeted driver alerts.

## 5. Demand Prediction Model (Random Forest)

```{r load-ml-data}
ml_df <- read.csv("data/processed/ml_dataset.csv", stringsAsFactors = FALSE)
ml_df$timestamp <- as.POSIXct(ml_df$timestamp, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")

FEATURES <- c("hour", "day_of_week", "is_weekend", "is_holiday",
              "temperature", "rainfall", "current_occupancy", "capacity", "average_duration")
TARGET <- "future_occupancy"

nrow(ml_df)
```

```{r train-model}
set.seed(42)
n <- nrow(ml_df)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))

train_df <- ml_df[train_idx, c(FEATURES, TARGET)]
test_df  <- ml_df[-train_idx, c(FEATURES, TARGET)]

model_formula <- as.formula(paste(TARGET, "~", paste(FEATURES, collapse = " + ")))

rf_model <- randomForest(
  model_formula, data = train_df,
  ntree = 200, mtry = max(1, floor(length(FEATURES) / 3)),
  importance = TRUE
)

preds <- predict(rf_model, newdata = test_df)

mae  <- mean(abs(preds - test_df[[TARGET]]))
rmse <- sqrt(mean((preds - test_df[[TARGET]])^2))
r2   <- 1 - sum((test_df[[TARGET]] - preds)^2) / sum((test_df[[TARGET]] - mean(test_df[[TARGET]]))^2)

data.frame(metric = c("MAE", "RMSE", "R2"), value = round(c(mae, rmse, r2), 4))
```

```{r feature-importance-plot, fig.cap="Feature importance (IncNodePurity)"}
imp <- importance(rf_model)
imp_df <- data.frame(feature = rownames(imp), importance = imp[, "IncNodePurity"]) %>%
  arrange(desc(importance))

ggplot(imp_df, aes(x = reorder(feature, importance), y = importance)) +
  geom_col(fill = "#31a354") +
  coord_flip() +
  labs(title = "Feature Importance", x = "Feature", y = "Increase in Node Purity") +
  theme_minimal()
```

As with the Python/scikit-learn version of this model (see `ml/train_model.py`), `current_occupancy` dominates feature importance — expected at a 30-minute prediction horizon, since occupancy is highly autocorrelated over such a short window. `hour` picks up most of the remaining signal, reflecting the peak-hour structure seen in Section 4.

## 6. Conclusion

- Zones show clear, reproducible peak/off-peak utilization patterns tied to time of day.
- A Random Forest trained on nine straightforward features predicts 30-minute-ahead occupancy with low error, primarily by extrapolating from current occupancy and time of day.
- The same modeling approach was implemented independently in Python (`ml/train_model.py`) and R (`r-analytics/train_model.R`) against the same dataset, as a basic cross-language sanity check on the results.

**Note:** update the exact MAE/RMSE/R² figures above with the numbers your own knit run produces before including this report in a submission — they depend on the dataset generation seed and parameters.
